Equations & R Code

### 2016-10-12 manuscript/coexistence-stability_ms.tex
updating results text

---
layout: 12pt
header-includes:
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{setspace}
  - \usepackage{todonotes}
  - \doublespacing
  - \usepackage{rotating}
  - \usepackage{color, soul}
  - \usepackage[font={footnotesize},labelfont={sf,bf},labelsep=space]{caption}
  - \usepackage{sectsty}
bibliography: ~/Dropbox/Bibliography/CoexistStability.bib
csl: components/ecology.csl

## rmarkdown render options
output:
  pdf_document:
    fig_caption: true
    keep_tex: true
fontsize: 12pt
geometry: margin=1in
linkcolor: black
urlcolor: black
---

\renewcommand\linenumberfont{\normalfont\tiny\sffamily\color{gray}}
\allsectionsfont{\normalfont\sffamily\bfseries}

@@

```{r caching, include=FALSE}
library("methods")
library("knitr")
basename <- "manuscript"
opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
               cache.path = paste("components/cache/", basename, "/", sep=""))
opts_chunk$set(cache = 2)
opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, 
               comment = NA, verbose = TRUE, echo=FALSE)

# PDF-based figures
opts_chunk$set(dev='pdf')
```

```{r plot-options, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

opts_chunk$set(dev='pdf', dev.args=c(version="1.3"))

```

@@

INTRODUCTION
------------
@MacArthur1955, @Elton1958, and even Darwin [@Turnbull2013] recognized that species can compensate for each other and stabilize functioning in ecosystems subject to temporal variation in environmental conditions. 
This idea underlies the "insurance hypothesis" [@Yachi1999], which suggests stability increases with diversity because species respond dissimilarly to environmental conditions – species A has highest growth rates under conditions X whereas species B has highest growth rates under conditions Y. 
More species confer temporal stability by broadening the range of conditions under which the community maintains function [@Loreau2010].
Diverse models all predict a positive relationship between species richness and ecosystem stability [@Lehman2000; @Ives2002b; @Loreau2013], and experimental tests tend to support such a prediction [@Tilman2006; @Hector2010].
However, empirical support for a positive diversity-stability relationship is not unequivocal [@Jiang2009].
Furthermore, theory on the relationship between biodiversity and ecosystem stability often ignores the processes that determine species coexistence in variable environments [Loreau 2010, but see @Chesson2001].

Temporally fluctuating environmental conditions are an important ingredient for stable species coexistence, both in theoretical models [@Chesson2000; @Chesson2004] and in natural communities [@Caceres1997; @Descamps-Julien2005; @Adler2006; @Angert2009].
Such "fluctuation-dependent" coexistence [@Chesson2000] requires that species have unique environmental responses and that environmental conditions vary enough for each coexisting species to experience good and bad conditions.
Thus, there is reason to expect environmental variability to promote species richness when coexistence is maintained by a fluctuation-dependent mechanism [@Adler2008].
Of course, increasing environmental variability may also decrease ecosystem stability through time by increasing the fluctuations of individual species, regardless of species richness.

The countervailing effects of environmental variability present an interesting paradox: increasing variability should decrease ecosystem stability, but may also increase richness, which may offset the decrease in stability.
Such a paradox complicates predictions about how ecosystems will respond as environmental conditions exceed historical ranges of variability.
The unknown net effect of environment variability may be reflected in the mixed results from empirical studies on the diversity-stability relationship.
Observational tests of the diversity-stability relationship, which require sampling across natural diversity gradients, have yielded positive [@Hautier2014], neutral [@Valone2003; @Cusson2015], and negative [@Sasaki2011] relationships.
In a meta-analysis of diversity-stability relationships, @Jiang2009 found no significant evidence for an effect of species richness on ecosystem stability from observational studies in terrestrial ecosystems.
Thus, there appears to be a gap between the consistency of theoretical studies and the equivocation of empirical studies.

We argue this gap exists because the two bodies of theory that have developed to explain species coexistence on the one hand, and diversity-stability relationships on the other, have diverged.
One reason these two disciplines have diverged is because they have focused on slightly different questions.
Biodiversity-ecosystem stability studies typically ask how ecosystem variability responds to different levels of species richness at a given level of environmental variability [reviewed in @Kinzig2001; @Loreau2010], whereas coexistence studies ask how the long term stability of species coexistence responds to different levels of environmental variability [@Chesson1981a].

To reconcile these two bodies of theory, we extend theory on the relationship between species richness and ecosystem stability to cases in which species coexistence explicitly depends on environmental fluctuations and species-specific responses to environmental conditions.
We focus on the storage effect and relative nonlinearity using a general consumer-resource model.
First, we use model simulations to investigate the diversity-stability relationship across a gradient of environmental variability.
Counter to common expectations, we find that a negative diversity-stability relationship should be expected when sampling occurs over natural diversity gradients and species coexistence is fluctuation dependent.
Importantly, and in line with previous theory [@Ives2002b; @Loreau2010; @DeMazancourt2013], at a given level of environmental variability, increasing species richness stabilizes ecosystem function, even when coexistence is fluctuation dependent.

@@

MATERIALS AND METHODS
---------------------
###  Consumer-resource model
We developed a semi-discrete consumer-resource model that allows many species to coexist on one resource by either the storage effect or relative nonlinearity. 
In our model, the consumer can be in one of two-states: a dormant state $D$ and a live state $N$.
The dormant state could represent, for example, the seedbank of an annual plant.
Transitions between $N$ and $D$ occur at discrete intervals $\tau$ with continuous-time consumer-resource dynamics between discrete transitions. 
Thus, our model is formulated as "pulsed differential equations" [@Pachepsky2008; @Mailleret2009; @Mordecai2016]. 
For clarity we refer to $\tau$ as years and the growing time between years as seasons with daily ($t$) time steps.

@@

\noindent where the discrete transitions between $N$ and $D$ occur between seasons at times $\tau_k$, $k = 1,2,3, \dots, K$. The subscript $i$ denotes species, $N$ is the living biomass state, and $\epsilon_i$ is each species' resource-to-biomass conversion efficiency. The growth rate of living biomass is a resource-dependent Hill function, $f_{i}(R) = r_{i}R^{a_{i}} / (b_{i}^{a_{i}}+R^{a_{i}})$, where *r* is a species' intrinsic growth rate and *a* and *b* define the curvature of the function. Resource depletion is equal to the sum of each species' consumption.

Along with resource uptake, consumer population growth depends on the production of dormant biomass ($D$), the activation of dormant biomass to live biomass ($D \rightarrow N$), and the survival of living biomass from one year to the next. The biomass of each species' states at the start of a growing season are equal to

@@

\noindent where $D(\tau_k)$, $N(\tau_k)$, and $R(\tau_k)$ are the abundances of each state at the end of growing season *k* and $\tau_k^+$ denotes the beginning of growing season $k=1$. 
The activation of dormant biomass to live biomass is controlled by $\gamma$, which is year (*k*) and species (*i*) specific.
Dormant biomass is equal to a constant fraction ($\alpha$) of live biomass at the end of the previous season ($N_{i}(\tau_k)$), plus survival ($1-\eta_i$) of dormant biomass ($D_{i}(\tau_k)$) at the end of the previous year and dormant biomass remaining after live biomass activation ($D_{i}(\tau_k)(1-\gamma_{i,\tau_k})$).
Live biomass is equal to newly activated dormant biomass ($\gamma_{i,t}[D_{i}(\tau_k)$), minus some fraction of live biomass that is converted to dormant biomass ($(1-\alpha_i)N_{i}(\tau_k)$)
We assume the resource pool is not replenished within a growing season.
Resource replenishment occurs between growing seasons, and the resource pool (*R*) at the start of the growing season *k*+1 is $R(\tau_k^+) = R^+$, where $R^+$ is a random resource pulse drawn from a log-normal distribution with mean $\mu(R^+)$ and standard deviation $\sigma(R^+)$. 
Model parameters and notation are described in Table 1.

#### Implementing the Storage Effect
To make this a storage effect model, we need to satisfy three conditions: (1) the organisms must have a mechanism for persistence under unfavorable conditions, (2) species must respond differently to environmental conditions, and (3) the effects of competition on a species must be more strongly negative in good years relative to unfavorable years. 
Our model meets condition 1 because we include a dormant stage with very low death rates. 
We satisfy condition 2 with our model whenever $\gamma$ is not perfectly correlated between species. 
Lastly, our model meets condition 3 because condition 2 partitions intraspecific and interspecific competition into different years. 
Thus, during a high $\gamma$ year for one species, resource uptake is also inherently high for that species, which increases intraspecific competition relative to interspecific competition. 
So, given adequate variability in $\gamma$, the inferior competitor can persist. 
We created competitive hierarchies in the storage effect version of the model by altering species' biomass conversion efficiencies ($\epsilon$)

We generated sequences of (un)correlated dormant-to-live state transition rates ($\gamma$) for each species by drawing from multivariate normal distributions with mean 0 and a variance-covariance matrix ($\Sigma(\gamma)$) of

@@

\noindent where $\sigma^2_{E}$ is the variance of the environmental cue and $\rho_{i,j}$ is the correlation between between the species *i*'s and species *j*'s transition rates. 
$\rho$ must be less than 1 for stable coexistence, and in all simulations we placed that constraint that $\rho_{i,j} = \rho_{j,i}$ for each species pair.
The inferior competitor has the strongest potential to persist when $\rho=-1$ (perfectly uncorrelated transition rates). 
We used the ```R``` function ```mvrnorm``` to generate sequences of (un)correlated variates *E* that we converted to germination rates in the 0-1 range: $\gamma = e^E / 1 + e^E$.
Note that $\Sigma(\gamma)$ must be positive definite.
So, after defining $\Sigma(\gamma)$ with all $\rho_{i,j}$s and $\sigma^2_{E}$, we used the ```nearPD``` function from the ```Matrix``` package in ```R``` to coerce the variance-covariance matrix to be positive definite.

#### Implementing Relative Nonlinearity
When considering consumer-resource dynamics, species coexistence by relative nonlinearity requires that each species has different nonlinear responses to resource availability, and resource availability must fluctuate through time.
In a constant resource environment, the species with the lowest $R^*$ will always exclude the other species. 
To create competitive hierarchies among species we manipulated species resource uptake curves such that the species with the lowest $R^*$ also had the lowest maximum growth rate at high resource levels (e.g., low *r*, low *a*, and low *b* values relative to other species in the Hill equation; Fig. SX).
Thus, our simulated species represent a continuum from resource-conservative to resource-acquisitive.
We still allow the germination rate ($\gamma$) to vary, but both species are perfectly correlated -- that is, $\rho=1$ (Fig. 1).

### Numerical simulations
To understand how fluctuation-dependent coexist can affect the diversity-stability relationship, we simulated the model with four species under two scenarios for each coexistence mechanism.
First, we allowed the variance of the environment to determine how many species can coexist, akin to a community assembly experiment with a species pool of four species.
This required simulating communities with all species initially present across a gradient of annual resource variability (for relative nonlinearity) or environmental cue variability (for the storage effect).
Second, we chose parameter values that allowed coexistence of all four species and performed species removals.
The two simulation experiments correspond to (i) sampling ecosystem function across a natural gradient of species richness and (ii) sampling ecosystem function across diversity treatments within a site.

@@

All simulations were run for 5,000 seasons with 20-day growing seasons. 
We averaged biomass over the growing season, and those yearly values were used to calculate total community biomass in each year.
After discarding an initial 500 seasons to reduce transient effects on our results, we calculated the coffecient of variation (*CV*) of summed species biomass through time.
Therefore, in our results we refer to ecosystem variability, which is the inverse of ecosystem stability.
We calculated species richness as the number of species whose average biomass was greater than 1 over the course of the simulation.
Parameter values for specific results are given in figure captions.
Within-season dynamics were solved given initial conditions using the package ``deSolve`` [@Soetaert2010] in ``R`` [@Team2013].
All model code has been deposited on Figshare (*link*) and is available on GitHub at http://github.com/atredennick/Coexistence-Stability.

RESULTS AND DISCUSSION
----------------------
### The diversity-variability relationship
The direction of the diversity-variability relationship can be positive and negative when species coexistence is maintained by fluctuation-dependent mechanisms (Fig. 2).
Ecosystem variability is positively correlated with species richness when species richness is measured across a gradient of environmental variability, which maintains diversity and promotes ecosystem variability (Fig. 2A,C).
If environmental conditions are sufficient to maintain coexistence, removing species increases ecosystem variability (Fig. 2B,D).
Thus, our results both confirm and contrast with theoretical and experimental findings that diversity begets stability.

@@

When we allowed a gradient of environmental variability to determine species coexistence, we discovered a positive relationship between species richness and ecosystem variability (Fig. 2A,C).
While surprising when viewed through the lens of previous theory and experimental findings, such a relationship is a direct consequence of how diversity can be maintained in fluctuating environments.
The storage effect and relative nonlinearity both require envrionmental fluctuations to allow niche differentiation between species pairs [@Chesson2000].
Therefore, species coexistence gains strength, for both mechanisms, as the environment becomes more variable (Fig. SX).

Our results may explain why deviations from the negative diversity-variability relationship often come from observational studies [@Jiang2009].
Observational studies must rely on natural diversity gradients, and if species richness depends on environmental variability, it is entirely possible to observe positive diversity-variability relationships.
For example, Sasaki and Lauenroth (2011) found a negative relationship between species richness and the temporal stability of plant abundance (a positive diversity-variability relationship) in a semi-arid grassland.
Their data came from a six sites that were 6 km apart.
While Sasaki and Lauenroth explained their results in terms of dominant species' effects, it is also possible that each site experienced slightly different levels of environmental variability that influenced species coexistence.
@DeClerck2006 also found a positive diversity-variability when sampling conifer richness and the variability of productivity across a large spatial gradient in the Sierra Nevada.

While our modeling results show that fluctuation-dependent coexistence can create positive diversity-variability relationships, whether such trends are detected will depend on the particular traits of the species in the community and the relative influence of fluctuation-dependent and fluctuation-independent coexistence mechanisms, which are not mutually exclusive.
Thus, our results may also help explain observational studies where no relationship between diversity and variability is detected.
For example, @Cusson2015 found no relationship between species richness and variability of abundances in several marine macro-benthic ecosystems.
Many of their focal ecosystems were from highly variable intertidal environments.
If coexistence was at least in part determined by environmental fluctuations, then the confounding effect of variability and species richness could compensate any direct effect of species richness on variability.
Previous theoretical work showed how environmental variation can mask the effect of species diversity on ecosystem productivity when sampling across sites [@Loreau1998].
Our mechanistic model extends that conclusion to ecosystem stability.

### The impact of increasing environmental variability on ecosystem variability
Whether coexistence is fluctuation-independent or fluctuation-dependent becomes especially important when we consider how ecosystem stability responds to increasing environmental variability.
In the fluctuation-independent case, species richness is essentially fixed because the species' inequalities that determine coexistence (niche and fitness differences) are not linked to environmental variability.
Therefore, increasing environmental variability will always increase ecosystem variability by increasing the fluctuations of individual species' abundances.

When species coexistence is fluctuation-dependent via the storage effect, increasing environmental variability has much more interesting effects on ecosystem variability.
In Fig. 3 we show storage effect simulation results where environmental variability determines species coexistence from a regional species pool of four species.
We also show results from nested subsets of the four species pool (e.g., only two species in the pool instead of four) to show the trajectory of ecosystem variability if new species are not present to join the local community.
In accordance with coexistence theory, we find that realized species richness increases with environmental variability and, in some cases, increasing variability can actually completely temper the effect of increasing environmental variability.
More species rich communities are less variable on average (e.g., lower intercepts in log-log space; Fig. SX) and increase in ecosystem *CV* at a slower rate (e.g., lower slopes in log-log space; Fig. SX).

The dampening effect of fluctuation-dependent coexistence on increasing environmental variability depends on the specific traits (parameter values) of the species in the regional pool.
Asymmetric competition makes it more difficult for new species to enter the local community (Fig. 3; compare top and bottom panels).
Relatively high competition also increases the rate at which ecosystem *CV* increases with environmental variance (Fig. SX).
This is because the abundance of inferior competitors is reduced and therefore does not influence ecosystem *CV* as much as when competition is symmetric.
The correlation of species' environmental responses also mediates the relationship between environmental variance, species richness, and ecosystem *CV*: lower correlations make it easier for new species to coexist and unique environmental responses are always stabilizing (Fig. 3).

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=black,
            linkcolor=black,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{}
  \pretitle{\vspace{\droptitle}}
  \posttitle{}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\usepackage{todonotes}
\doublespacing
\usepackage{rotating}
\usepackage{color, soul}
\usepackage[font={footnotesize},labelfont={sf,bf},labelsep=space]{caption}
\usepackage{sectsty}

\begin{document}

\maketitle

\renewcommand\linenumberfont{\normalfont\tiny\sffamily\color{gray}}

\allsectionsfont{\normalfont\sffamily\bfseries}

@@

@@

\subsection{INTRODUCTION}\label{introduction}

MacArthur (1955), Elton (1958), and even Darwin (Turnbull et al. 2013)
recognized that species can compensate for each other and stabilize
functioning in ecosystems subject to temporal variation in environmental
conditions. This idea underlies the ``insurance hypothesis'' (Yachi and
Loreau 1999), which suggests stability increases with diversity because
species respond dissimilarly to environmental conditions -- species A
has highest growth rates under conditions X whereas species B has
highest growth rates under conditions Y. More species confer temporal
stability by broadening the range of conditions under which the
community maintains function (Loreau 2010). Diverse models all predict a
positive relationship between species richness and ecosystem stability
(Lehman and Tilman 2000, Ives and Hughes 2002, Loreau and {{de
Mazancourt}} 2013), and experimental tests tend to support such a
prediction (Tilman et al. 2006, Hector et al. 2010). However, empirical
support for a positive diversity-stability relationship is not
unequivocal (Jiang and Pu 2009). Furthermore, theory on the relationship
between biodiversity and ecosystem stability often ignores the processes
that determine species coexistence in variable environments (Loreau
2010, but see Chesson et al. 2001).

Temporally fluctuating environmental conditions are an important
ingredient for stable species coexistence, both in theoretical models
(Chesson 2000, Chesson et al. 2004) and in natural communities
(C{á}ceres 1997, Descamps-Julien and Gonzalez 2005, Adler et al. 2006,
Angert et al. 2009). Such ``fluctuation-dependent'' coexistence (Chesson
2000) requires that species have unique environmental responses and that
environmental conditions vary enough for each coexisting species to
experience good and bad conditions. Thus, there is reason to expect
environmental variability to promote species richness when coexistence
is maintained by a fluctuation-dependent mechanism (Adler and Drake
2008). Of course, increasing environmental variability may also decrease
ecosystem stability through time by increasing the fluctuations of
individual species, regardless of species richness.

The countervailing effects of environmental variability present an
interesting paradox: increasing variability should decrease ecosystem
stability, but may also increase richness, which may offset the decrease
in stability. Such a paradox complicates predictions about how
ecosystems will respond as environmental conditions exceed historical
ranges of variability. The unknown net effect of environment variability
may be reflected in the mixed results from empirical studies on the
diversity-stability relationship. Observational tests of the
diversity-stability relationship, which require sampling across natural
diversity gradients, have yielded positive (Hautier et al. 2014),
neutral (Valone and Hoffman 2003, Cusson et al. 2015), and negative
(Sasaki and Lauenroth 2011) relationships. In a meta-analysis of
diversity-stability relationships, Jiang and Pu (2009) found no
significant evidence for an effect of species richness on ecosystem
stability from observational studies in terrestrial ecosystems. Thus,
there appears to be a gap between the consistency of theoretical studies
and the equivocation of empirical studies.

We argue this gap exists because the two bodies of theory that have
developed to explain species coexistence on the one hand, and
diversity-stability relationships on the other, have diverged. One
reason these two disciplines have diverged is because they have focused
on slightly different questions. Biodiversity-ecosystem stability
studies typically ask how ecosystem variability responds to different
levels of species richness at a given level of environmental variability
(reviewed in Kinzig et al. 2001, Loreau 2010), whereas coexistence
studies ask how the long term stability of species coexistence responds
to different levels of environmental variability (Chesson and Warner
1981).

To reconcile these two bodies of theory, we extend theory on the
relationship between species richness and ecosystem stability to cases
in which species coexistence explicitly depends on environmental
fluctuations and species-specific responses to environmental conditions.
We focus on the storage effect and relative nonlinearity using a general
consumer-resource model. First, we use model simulations to investigate
the diversity-stability relationship across a gradient of environmental
variability. Counter to common expectations, we find that a negative
diversity-stability relationship should be expected when sampling occurs
over natural diversity gradients and species coexistence is fluctuation
dependent. Importantly, and in line with previous theory (Ives and
Hughes 2002, Loreau 2010, Mazancourt et al. 2013), at a given level of
environmental variability, increasing species richness stabilizes
ecosystem function, even when coexistence is fluctuation dependent.

@@

\subsection{MATERIALS AND METHODS}\label{materials-and-methods}

\subsubsection{Consumer-resource model}\label{consumer-resource-model}

We developed a semi-discrete consumer-resource model that allows many
species to coexist on one resource by either the storage effect or
relative nonlinearity. In our model, the consumer can be in one of
two-states: a dormant state \(D\) and a live state \(N\). The dormant
state could represent, for example, the seedbank of an annual plant.
Transitions between \(N\) and \(D\) occur at discrete intervals \(\tau\)
with continuous-time consumer-resource dynamics between discrete
transitions. Thus, our model is formulated as ``pulsed differential
equations'' (Pachepsky et al. 2008, Mailleret and Lemesle 2009, Mordecai
et al. 2016). For clarity we refer to \(\tau\) as years and the growing
time between years as seasons with daily (\(t\)) time steps.

@@

\noindent where the discrete transitions between \(N\) and \(D\) occur
between seasons at times \(\tau_k\), \(k = 1,2,3, \dots, K\). The
subscript \(i\) denotes species, \(N\) is the living biomass state, and
\(\epsilon_i\) is each species' resource-to-biomass conversion
efficiency. The growth rate of living biomass is a resource-dependent
Hill function,
\(f_{i}(R) = r_{i}R^{a_{i}} / (b_{i}^{a_{i}}+R^{a_{i}})\), where
\emph{r} is a species' intrinsic growth rate and \emph{a} and \emph{b}
define the curvature of the function. Resource depletion is equal to the
sum of each species' consumption.

Along with resource uptake, consumer population growth depends on the
production of dormant biomass (\(D\)), the activation of dormant biomass
to live biomass (\(D \rightarrow N\)), and the survival of living
biomass from one year to the next. The biomass of each species' states
at the start of a growing season are equal to

@@

\noindent where \(D(\tau_k)\), \(N(\tau_k)\), and \(R(\tau_k)\) are the
abundances of each state at the end of growing season \emph{k} and
\(\tau_k^+\) denotes the beginning of growing season \(k=1\). The
activation of dormant biomass to live biomass is controlled by
\(\gamma\), which is year (\emph{k}) and species (\emph{i}) specific.
Dormant biomass is equal to a constant fraction (\(\alpha\)) of live
biomass at the end of the previous season (\(N_{i}(\tau_k)\)), plus
survival (\(1-\eta_i\)) of dormant biomass (\(D_{i}(\tau_k)\)) at the
end of the previous year and dormant biomass remaining after live
biomass activation (\(D_{i}(\tau_k)(1-\gamma_{i,\tau_k})\)). Live
biomass is equal to newly activated dormant biomass
(\(\gamma_{i,t}[D_{i}(\tau_k)\)), minus some fraction of live biomass
that is converted to dormant biomass (\((1-\alpha_i)N_{i}(\tau_k)\)) We
assume the resource pool is not replenished within a growing season.
Resource replenishment occurs between growing seasons, and the resource
pool (\emph{R}) at the start of the growing season \emph{k}+1 is
\(R(\tau_k^+) = R^+\), where \(R^+\) is a random resource pulse drawn
from a log-normal distribution with mean \(\mu(R^+)\) and standard
deviation \(\sigma(R^+)\). Model parameters and notation are described
in Table 1.

\paragraph{Implementing the Storage
Effect}\label{implementing-the-storage-effect}

To make this a storage effect model, we need to satisfy three
conditions: (1) the organisms must have a mechanism for persistence
under unfavorable conditions, (2) species must respond differently to
environmental conditions, and (3) the effects of competition on a
species must be more strongly negative in good years relative to
unfavorable years. Our model meets condition 1 because we include a
dormant stage with very low death rates. We satisfy condition 2 with our
model whenever \(\gamma\) is not perfectly correlated between species.
Lastly, our model meets condition 3 because condition 2 partitions
intraspecific and interspecific competition into different years. Thus,
during a high \(\gamma\) year for one species, resource uptake is also
inherently high for that species, which increases intraspecific
competition relative to interspecific competition. So, given adequate
variability in \(\gamma\), the inferior competitor can persist. We
created competitive hierarchies in the storage effect version of the
model by altering species' biomass conversion efficiencies
(\(\epsilon\))

We generated sequences of (un)correlated dormant-to-live state
transition rates (\(\gamma\)) for each species by drawing from
multivariate normal distributions with mean 0 and a variance-covariance
matrix (\(\Sigma(\gamma)\)) of

@@

\noindent where \(\sigma^2_{E}\) is the variance of the environmental
cue and \(\rho_{i,j}\) is the correlation between between the species
\emph{i}'s and species \emph{j}'s transition rates. \(\rho\) must be
less than 1 for stable coexistence, and in all simulations we placed
that constraint that \(\rho_{i,j} = \rho_{j,i}\) for each species pair.
The inferior competitor has the strongest potential to persist when
\(\rho=-1\) (perfectly uncorrelated transition rates). We used the
\texttt{R} function \texttt{mvrnorm} to generate sequences of
(un)correlated variates \emph{E} that we converted to germination rates
in the 0-1 range: \(\gamma = e^E / 1 + e^E\). Note that
\(\Sigma(\gamma)\) must be positive definite. So, after defining
\(\Sigma(\gamma)\) with all \(\rho_{i,j}\)s and \(\sigma^2_{E}\), we
used the \texttt{nearPD} function from the \texttt{Matrix} package in
\texttt{R} to coerce the variance-covariance matrix to be positive
definite.

\paragraph{Implementing Relative
Nonlinearity}\label{implementing-relative-nonlinearity}

When considering consumer-resource dynamics, species coexistence by
relative nonlinearity requires that each species has different nonlinear
responses to resource availability, and resource availability must
fluctuate through time. In a constant resource environment, the species
with the lowest \(R^*\) will always exclude the other species. To create
competitive hierarchies among species we manipulated species resource
uptake curves such that the species with the lowest \(R^*\) also had the
lowest maximum growth rate at high resource levels (e.g., low \emph{r},
low \emph{a}, and low \emph{b} values relative to other species in the
Hill equation; Fig. SX). Thus, our simulated species represent a
continuum from resource-conservative to resource-acquisitive. We still
allow the germination rate (\(\gamma\)) to vary, but both species are
perfectly correlated -- that is, \(\rho=1\) (Fig. 1).

\subsubsection{Numerical simulations}\label{numerical-simulations}

To understand how fluctuation-dependent coexist can affect the
diversity-stability relationship, we simulated the model with four
species under two scenarios for each coexistence mechanism. First, we
allowed the variance of the environment to determine how many species
can coexist, akin to a community assembly experiment with a species pool
of four species. This required simulating communities with all species
initially present across a gradient of annual resource variability (for
relative nonlinearity) or environmental cue variability (for the storage
effect). Second, we chose parameter values that allowed coexistence of
all four species and performed species removals. The two simulation
experiments correspond to (i) sampling ecosystem function across a
natural gradient of species richness and (ii) sampling ecosystem
function across diversity treatments within a site.

@@

All simulations were run for 5,000 seasons with 20-day growing seasons.
We averaged biomass over the growing season, and those yearly values
were used to calculate total community biomass in each year. After
discarding an initial 500 seasons to reduce transient effects on our
results, we calculated the coffecient of variation (\emph{CV}) of summed
species biomass through time. Therefore, in our results we refer to
ecosystem variability, which is the inverse of ecosystem stability. We
calculated species richness as the number of species whose average
biomass was greater than 1 over the course of the simulation. Parameter
values for specific results are given in figure captions. Within-season
dynamics were solved given initial conditions using the package
\texttt{deSolve} (Soetaert et al. 2010) in \texttt{R} (Team 2013). All
model code has been deposited on Figshare (\emph{link}) and is available
on GitHub at \url{http://github.com/atredennick/Coexistence-Stability}.

\subsection{RESULTS AND DISCUSSION}\label{results-and-discussion}

\subsubsection{The diversity-variability
relationship}\label{the-diversity-variability-relationship}

The direction of the diversity-variability relationship can be positive
and negative when species coexistence is maintained by
fluctuation-dependent mechanisms (Fig. 2). Ecosystem variability is
positively correlated with species richness when species richness is
measured across a gradient of environmental variability, which maintains
diversity and promotes ecosystem variability (Fig. 2A,C). If
environmental conditions are sufficient to maintain coexistence,
removing species increases ecosystem variability (Fig. 2B,D). Thus, our
results both confirm and contrast with theoretical and experimental
findings that diversity begets stability.

@@

When we allowed a gradient of environmental variability to determine
species coexistence, we discovered a positive relationship between
species richness and ecosystem variability (Fig. 2A,C). While surprising
when viewed through the lens of previous theory and experimental
findings, such a relationship is a direct consequence of how diversity
can be maintained in fluctuating environments. The storage effect and
relative nonlinearity both require envrionmental fluctuations to allow
niche differentiation between species pairs (Chesson 2000). Therefore,
species coexistence gains strength, for both mechanisms, as the
environment becomes more variable (Fig. SX).

Our results may explain why deviations from the negative
diversity-variability relationship often come from observational studies
(Jiang and Pu 2009). Observational studies must rely on natural
diversity gradients, and if species richness depends on environmental
variability, it is entirely possible to observe positive
diversity-variability relationships. For example, Sasaki and Lauenroth
(2011) found a negative relationship between species richness and the
temporal stability of plant abundance (a positive diversity-variability
relationship) in a semi-arid grassland. Their data came from a six sites
that were 6 km apart. While Sasaki and Lauenroth explained their results
in terms of dominant species' effects, it is also possible that each
site experienced slightly different levels of environmental variability
that influenced species coexistence. DeClerck et al. (2006) also found a
positive diversity-variability when sampling conifer richness and the
variability of productivity across a large spatial gradient in the
Sierra Nevada.

While our modeling results show that fluctuation-dependent coexistence
can create positive diversity-variability relationships, whether such
trends are detected will depend on the particular traits of the species
in the community and the relative influence of fluctuation-dependent and
fluctuation-independent coexistence mechanisms, which are not mutually
exclusive. Thus, our results may also help explain observational studies
where no relationship between diversity and variability is detected. For
example, Cusson et al. (2015) found no relationship between species
richness and variability of abundances in several marine macro-benthic
ecosystems. Many of their focal ecosystems were from highly variable
intertidal environments. If coexistence was at least in part determined
by environmental fluctuations, then the confounding effect of
variability and species richness could compensate any direct effect of
species richness on variability. Previous theoretical work showed how
environmental variation can mask the effect of species diversity on
ecosystem productivity when sampling across sites (Loreau 1998). Our
mechanistic model extends that conclusion to ecosystem stability.

\subsubsection{The impact of increasing environmental variability on
ecosystem
variability}\label{the-impact-of-increasing-environmental-variability-on-ecosystem-variability}

Whether coexistence is fluctuation-independent or fluctuation-dependent
becomes especially important when we consider how ecosystem stability
responds to increasing environmental variability. In the
fluctuation-independent case, species richness is essentially fixed
because the species' inequalities that determine coexistence (niche and
fitness differences) are not linked to environmental variability.
Therefore, increasing environmental variability will always increase
ecosystem variability by increasing the fluctuations of individual
species' abundances.

When species coexistence is fluctuation-dependent via the storage
effect, increasing environmental variability has much more interesting
effects on ecosystem variability. In Fig. 3 we show storage effect
simulation results where environmental variability determines species
coexistence from a regional species pool of four species. We also show
results from nested subsets of the four species pool (e.g., only two
species in the pool instead of four) to show the trajectory of ecosystem
variability if new species are not present to join the local community.
In accordance with coexistence theory, we find that realized species
richness increases with environmental variability and, in some cases,
increasing variability can actually completely temper the effect of
increasing environmental variability. More species rich communities are
less variable on average (e.g., lower intercepts in log-log space; Fig.
SX) and increase in ecosystem \emph{CV} at a slower rate (e.g., lower
slopes in log-log space; Fig. SX).
